Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-11 resubmission/04-analisis.qmd
Update carpeta resubmission y varios (archivos no renderizados)

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---

@@

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average appears for perceived economic inequality in 2016, although this variable shows a downward trend over time, with the exception of 2019, which can be associated with the “social crisis” in the country. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. The meritocracy measures remain stable, though the perception that individuals are rewarded for effort is slightly higher than the perception that they are rewarded for their intelligence

@@

@tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects. The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals. Complementary, 77% of the variation corresponds to within-individual differences over time.

@@

m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
                data = df_study1)

@@

m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
                data = df_study1)

m1.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
                data = df_study1)

m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta),
                data = df_study1)

@@

m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + (1 + ola_num | idencuesta), 
           data = df_study1)

m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + (1 + ola_num | idencuesta),
                data = df_study1)

m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  (1 + ola_num | idencuesta),
                data = df_study1)

m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
                data = df_study1)

m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
                data = df_study1)

# interactions 
m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc_dic*perc_inequality_cwc_dic + merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

m8.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

m8.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_mean*perc_inequality_mean + merit_talent_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1)

m9.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1)

m9.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1)

@@

m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality + ola_num| idencuesta),
                data = df_study1)

m10.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola_num + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
                data = df_study1)

m10.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola_num + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
                data = df_study1)

m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + merit_effort*ola_num  + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort + ola_num| idencuesta),
                data = df_study1)

m11.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola_num  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1)

m11.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola_num + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
                data = df_study1)

m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent + ola_num| idencuesta),
                data = df_study1)

m12.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1)

m12.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean*ola_num + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_mean + ola_num| idencuesta),
                data = df_study1)

m13 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality  + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort*perc_inequality*ola_num +
              (1 + perc_inequality + merit_effort  + ola_num| idencuesta),
                data = df_study1)

m13.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1)

m13.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             ola_num*merit_effort_mean*perc_inequality_mean +
              (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
                data = df_study1)

m14 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent*perc_inequality*ola_num +
              (1 + perc_inequality + merit_talent + ola_num| idencuesta),
                data = df_study1)

m14.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1)

m14.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_mean*perc_inequality_mean*ola_num +
              (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
                data = df_study1)

@@

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences. Model 4 shows that meritocratic perceptions operate in the opposite direction. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

@@

```{r}
#| label: fig-interact
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Between effects of meritocratic perceptions on market justice preferences by economic inequality perception"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

ggeffects::ggpredict(model = m9.1,
                     terms = c("merit_effort_mean",
                               "perc_inequality_mean [min, max]"),
                     ci_level = .95
                     ) %>% 
  plot()

predictions(m9.1,
            condition = c("merit_effort_mean", "perc_inequality_mean"),
            newdata = datagrid(merit_effort_mean = df_study1$merit_effort_mean,
                               perc_inequality_mean = seq(min(df_study1$perc_inequality_mean), max(df_study1$perc_inequality_mean), by= 0.1)),
            re.form = NA) %>% 
  ggplot(aes(y =estimate,x=merit_effort_mean, fill=perc_inequality_mean, color=perc_inequality_mean)) +
  geom_ribbon(alpha=0.8,size=1,linetype="solid",color="black") +
  geom_line(size=1,color="black") +
  geom_point(size=2,color="black")

p6 <- marginaleffects::plot_slopes(model = m9.1,
                             variables = "merit_effort_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Effort (BE)",
       fill = NULL,
       title = "(A) People are rewarded by their efforts") +
  theme_ggdist()

p7 <- marginaleffects::plot_slopes(model = m9.2,
                             variables = "merit_talent_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Talent (BE)",
       fill = NULL,
       title = "(B) People are rewarded for their intelligence",
       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4.
Confidence intervals at 95%") +
  theme_ggdist()

library(ggpubr)
ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.

>>>>

---
author: "Equipo EDUMER"
bibliography: "../input/bib/merit-edjust.bib"
csl: "../input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---

@@

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average values across survey waves. The results show a notable upward trend in market justice preferences, particularly in the most recent waves. Perceived economic inequality exhibited the highest standardized mean in 2016; however, this variable shows a general downward trajectory over time, with a temporary increase in 2019, possibly reflecting the effects of the social uprising that year. Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise. Meritocratic beliefs, in contrast, remain relatively stable overall. Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort. Both meritocracy-related measures follow a similar temporal pattern: an increase from 2016 to 2018, a decline between 2019 and 2022—potentially associated with the social unrest and the consequences of the COVID-19 pandemic—and a subsequent rise in 2023, returning to levels comparable to those observed in 2016.

@@

@tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects. The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.31, indicating that approximately 31% of the variation is attributable to differences between individuals. Complementary, 69% of the variation corresponds to within-individual differences over time.

@@

m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
                data = df_study1, weights = ponderador_long_total)

@@

m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m1.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta),
                data = df_study1, weights = ponderador_long_total)

@@

m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + (1 + ola_num | idencuesta), 
           data = df_study1, weights = ponderador_long_total)

m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + (1 + ola_num | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  (1 + ola_num | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
                data = df_study1, weights = ponderador_long_total)

# interactions 
m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc_dic*perc_inequality_cwc_dic + merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m8.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m8.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_mean*perc_inequality_mean + merit_talent_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m9.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1, weights = ponderador_long_total)

m9.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1, weights = ponderador_long_total)

@@

m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m10.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola_num + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m10.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola_num + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + merit_effort*ola_num  + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m11.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola_num  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m11.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola_num + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m12.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m12.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean*ola_num + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_mean + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m13 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality  + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort*perc_inequality*ola_num +
              (1 + perc_inequality + merit_effort  + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m13.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m13.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             ola_num*merit_effort_mean*perc_inequality_mean +
              (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m14 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent*perc_inequality*ola_num +
              (1 + perc_inequality + merit_talent + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m14.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

m14.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_mean*perc_inequality_mean*ola_num +
              (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
                data = df_study1, weights = ponderador_long_total)

@@

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.039 point decrease in market justice preferences. Model 4 shows that meritocratic perceptions operate in the opposite direction. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.027, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

@@

```{r}
#| label: fig-interact
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Predicted values of market justice preferences by perceptions of meritocraticy and economic inequality"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

min_val <- min(df_study1$perc_inequality_mean, na.rm = TRUE)
max_val <- max(df_study1$perc_inequality_mean, na.rm = TRUE)
mean_val <- mean(df_study1$perc_inequality_mean, na.rm = TRUE)

p6 <- ggpredict(
  m9.1,
  terms = c("merit_effort_mean", 
            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
) %>% 
  plot() +
  scale_x_continuous(n.breaks = 5) +
  scale_color_manual(
    name = "Perception of inequality (BE)",
    values = c("#0571B0", "black", "#B2182B"),
    labels = c("Min", "Mean", "Max")
  ) +
  scale_fill_manual(
    name = "Perception of inequality (BE)",
    values = c("#0571B0", "black", "#B2182B"),
    labels = c("Min", "Mean", "Max")) +
  labs(
    x = "Merit: Effort (BE)",
    y = "Market justice preferences",
    title = "(A) People are rewarded by their efforts") +
  theme_ggdist()

p7 <- ggpredict(
  m9.2,
  terms = c("merit_talent_mean", 
            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
) %>% 
  plot() +
  scale_x_continuous(n.breaks = 5) +
  scale_color_manual(
    name = "Perception of inequality (BE)",
    values = c("#0571B0", "black", "#B2182B"),
    labels = c("Min", "Mean", "Max")
  ) +
  scale_fill_manual(
    name = "Perception of inequality (BE)",
    values = c("#0571B0", "black", "#B2182B"),
    labels = c("Min", "Mean", "Max")) +
  labs(
    x = "Merit: Talent (BE)",
    y = "Market justice preferences",
    title = "(B) People are rewarded for their intelligence",
  caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4") +
  theme_ggdist()

library(ggpubr)
ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1, legend = "top")

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. Specifically, these positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. As shown in @fig-interact, Panel A, the effect of meritocratic perceptions about effort on market justice preferences strengthens as perceived inequality increases, particularly among individuals who perceive the highest levels of inequality, whereas this effect weakens among those who perceive low levels of inequality. Panel B reveals a similar pattern for meritocratic perceptions about talent, although the effect is substantially weaker at low levels of perceived inequality. Substantively, these findings suggest that the legitimizing power of meritocratic perceptions is amplified in contexts perceived as highly unequal, reinforcing individualistic understandings of justice and strengthening support for market-based access to essential social services.
